A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2024-11-14 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: more work on fraction capacity

Note that the mixing coefficient $p$ is connected to probability but it is not itself probability. It represents a parameter for the mixing states, mixing preparations, while probability is typically understood as the chance of something happening, the likelihood of a final event. Mathematically, the probability measure is a function from an event, which can be understood as a collection of fully distinct outcomes, to a number from zero to one. If we are mixing ensemble that represent fully distinct outcomes (i.e. they are orthogonal), then the mixing $p$ is a probability. If we are mixing ensemble that do not represent fully distinct outcomes (e.g. spin up and spin left), the probability of getting one of the outcomes is not the mixing coefficient $p$.

This is exactly what happens in quantum mechanics: probability is defined only on a set of orthogonal states. If we focus only on preparation, then, we only talk about mixing coefficients across non-orthogonal elements and define a more general tool that, when restricted on orthogonal sets, will recover probability.

We concentrate on the following question: given an ensemble $\ens \in \Ens$ and a (Borel) set of ensembles $A \subset \Ens$, what fraction of $\ens$ can be constructed by a mixture of $A$? How much of $\ens$ can be explained as coming from preparations corresponding to $A$? For example, if $\ens$ is a uniform distribution over a fair six-sided die, and $A$ contains four ensemble, each fully on ${1,2}$ and uniform distribution over ${5,6}$ two and four, than $5/6$ of $\ens$ can be constructed from $A$, which is the probability that we d

Given a target ensemble $\ens \in \Ens$ and an arbitrary ensemble $\ens[a] \in \Ens$ we will be able to write $\ens = p \ens[a] + (1-p) \ens[b]$ for some $p \in [0,1]$ and some $\ens[b] \in \Ens$. At the very least, we can use $p=0$ and $\ens[b] = \ens$. Given that $p$ is bounded, there will be a biggest $p$ possible which we call the \textbf{fraction} of $\ens[a]$ in $\ens$. Mathematically
\begin{equation}
	\fraction_{\ens}(\ens[a]) = \sup(\{ p \in [0,1] \, | \, \exists \, \ens_1 \in \Ens \text{ s.t. }  \ens = p \ens[a] + \bar{p} \ens_1 \}).
\end{equation}
This quantity tells us how much of ensemble $\ens$ can be constructed, can be characterized, by $\ens[a]$.

We now extend this idea from a single ensemble $\ens[a] \in \Ens$ to a set of ensembles $A \subset \Ens$., meaning we want to characterize how much of a target ensemble $\ens$ can be constructed as a mixture of ensembles from $A$. Given a set of ensembles $A$, its \textbf{hull}, noted $\hull(A)$ is the set of all mixtures that can be constructed from $A$. That is, is the set of all convex combinations $\sum_i^n p_i \ens[a]_i$ such that $\{a_i\}_1^n \subseteq A$. The \textbf{fraction capacity} is the biggest fraction among all the elements of the hull. Mathematically
\begin{equation}
	\frcap_{\ens}(A) = \sup(\fraction_{\ens}(\hull(A))\cup\{0\}).
\end{equation}

If $\Ens$ is a topological space, we can imagine the fraction capacity to be defined on the sigma algebra $\Sigma_{\Ens}$. We can then show that the fraction capacity $\frcap_{\ens} : \Sigma_{\Ens} \to [0,1]$ is a set function that satisfies the following:
\begin{enumerate}
	\item non-negative and unit bounded - $0 \leq \frcap_{\ens}(A) \leq 1$
	\item monotone - $A \subseteq B \implies \frcap_{\ens}(A) \leq \frcap_{\ens}(B)$
	\item sub-additive - $\frcap_{\ens}(A \cup B) \leq \frcap_{\ens}(A) + \frcap_{\ens}(B)$.
	\item continuous from below and above - $\frcap_{\ens}(\lim\limits_{i \to \infty} A_i) = \lim\limits_{i \to \infty} \frcap_{\ens}(A_i)$ for any increasing or decreasing sequence $\{A_i\}$.
\end{enumerate}

>>>>

Note that the mixing coefficient $p$ is connected to probability but it is not itself probability in the standard sense. It represents a parameter for the mixing states, mixing preparations, while probability is typically understood as the chance of something happening, the likelihood of a final event. Mathematically, the probability measure is a function from an event, which can be understood as a collection of fully distinct outcomes, to a number from zero to one. If we are mixing ensembles that represent fully distinct outcomes (i.e. they are orthogonal), then the mixing $p$ is a probability in the standard sense. If we are mixing ensembles that do not represent fully distinct outcomes, that are not orthogonal like spin up and spin left, the probability of getting one of the outcomes is not the mixing coefficient $p$.

We essentially have two ways in which two events can be compared. We have \textbf{disjoint events}, those that represent disjoint sets of different states, like spin up and spin down; we also have \textbf{orthogonal events}, those that represent represent orthogonal sets of states, like spin up and spin down. The standard notion of probability applies only when considering orthogonal events. Classical probability rests on the assumption that all disjoint events are orthogonal. That is, that all states can be perfectly distinguished. Quantum mechanics breaks this assumption, and therefore we need a probability theory that can deal with disjoint but non-orthogonal events.

Let us concentrate on the following question: given an ensemble $\ens \in \Ens$ and a (Borel) set of ensembles $A \subseteq \Ens$, what fraction of $\ens$ can be constructed by a mixture of $A$? How much of $\ens$ can be explained as coming from preparations corresponding to $A$? For example, let $\Ens$ be the space of all probability distributions for a six-sided die. Let $\ens_{123456}$ be the uniform distribution over all outcomes. Let $A_1 = \{\ens_1\}$ where $\ens_1$ represents outcome one with 100\% probability. Since we can write $\ens_{123456} = \frac{1}{6} \ens_{1} + \frac{5}{6} \ens_{23456}$, where $\ens_{23456}$ represents a uniform distributions over the five outcomes, $1/6$ of $\ens$ can be constructed from $A$, but no more. Similarly, if $A_{12} = \{\ens_{1},\ens_{2}\}$, we can write $\ens = \frac{1}{3} \left(\frac{1}{2} \ens_1 + \frac{1}{2} \ens_2 \right)  + \frac{2}{3} \ens_{3456}$, so $1/3$ of $\ens$ can be constructed from $A$, but no more. Note that, given the uniform distribution, $1/6$ is exactly the probably for the event $A_1$ and $1/3$ the probability for event $A_{12}$. This gives the basic insight for our definition.

Given a target ensemble $\ens \in \Ens$ and an arbitrary ensemble $\ens[a] \in \Ens$, we define the \textbf{fraction} of $\ens[a]$ in $\ens$ to be
\begin{equation}
	\fraction_{\ens}(\ens[a]) = \sup(\{ p \in [0,1] \, | \, \exists \, \ens_1 \in \Ens \text{ s.t. }  \ens = p \ens[a] + \bar{p} \ens_1 \}).
\end{equation}
The fraction is always well-defined because we can always write $\ens = 0 \ens[a] + 1 \ens$, therefore it must be zero or greater. This quantity tells us how much of ensemble $\ens$ can be constructed from $\ens[a]$.

We now extend this idea from a single ensemble $\ens[a] \in \Ens$ to a Borel set of ensembles $A \subset \Ens$. Recalling that the $\hull(A)$ is the set of all possible convex combinations (i.e. mixtures), we define the\textbf{fraction capacity} of $A$ for $\ens$ to be
\begin{equation}
	\frcap_{\ens}(A) = \sup(\fraction_{\ens}(\hull(A))\cup\{0\}).
\end{equation}
This returns the biggest fraction of $\ens$ that can be achieved with a mixture of elements of $A$.\footnote{The name fraction capacity is chosen to both signify the ability of the set $A$ to contain $\ens$ (i.e. fraction capacity of one means $\ens$ is within the convex combinations of $A$) and to indicate that it will be a non-additive monotonic measure, which are called ``capacities'' in some literature. TODO \href{ https://link.springer.com/book/10.1007/978-3-319-03155-2}{cite} }

One can then that the fraction capacity $\frcap_{\ens} : \Sigma_{\Ens} \to [0,1]$ is a set function that satisfies the following:
\begin{enumerate}
	\item non-negative and unit bounded - $0 \leq \frcap_{\ens}(A) \leq 1$
	\item monotone - $A \subseteq B \implies \frcap_{\ens}(A) \leq \frcap_{\ens}(B)$
	\item sub-additive - $\frcap_{\ens}(A \cup B) \leq \frcap_{\ens}(A) + \frcap_{\ens}(B)$
	\item continuous from below and above - $\frcap_{\ens}(\lim\limits_{i \to \infty} A_i) = \lim\limits_{i \to \infty} \frcap_{\ens}(A_i)$ for any increasing or decreasing sequence $\{A_i\}$.
\end{enumerate}

The first property readily comes from the domain of the fraction. For the second, note that $\hull$ and $\sup$ are both monotone. For sub-additivity, note that the fraction of two components can at most sum during mixing. For the last, increasing and decreasing sequences strictly add or remove possible mixtures, therefore they will lead to increasing or decreasing sequences of real numbers bounded between zero and one. The limit of these sequences must agree with the fraction capacity of the limit.

Note that a probability satisfies all the properties of a fraction capacity, but it also satisfy on additional property: additivity over disjoint sets. That is, $\mu(A \cup B) = \mu(A) + \mu(B)$ if $A \cup B = \emptyset$. This is not true in general, as we can see in a simple example from quantum mechanics. Let $\Ens$ be the ensemble space of a qubit, that is the space of density matrices of a two dimensional Hilbert space. Geometrically, this is the convex set represented by a three dimensional ball. Let $\ens_{\psi} = |\psi\>\<\psi|$ be a pure state and let $\ens_{\phi} = |\phi\>\<\phi|$ be its orthogonal. Geometrically, these are two opposite points on the surface of the ball. The maximally mixed state can be expressed as $\ens = \frac{1}{2} \ens_0 + \frac{1}{2} \ens_1$. This is the center point of the ball and is in fact the midpoint between two opposite points. Since this is valid for any point, we have $\frcap_{\ens}(\{\ens_{\psi}\})=\frac{1}{2}$ for any $\ens_{\psi} \in \Ens$ and $\frcap_{\ens}(\{\ens_{\psi}, \ens_{\phi}\})=1$ for any pair of orthogonal pure states. If $A$ is a set of two orthogonal pure states, and $B$ is state different from the two, we have $\frcap_{\ens}(A \cup B) = 1 < \frcap_{\ens}(A) + \frcap_{\ens}(B) = \frac{3}{2}$.

If we restrict ourselves to lattice of subsets of $\Ens$ we can recover additivity. For example, if we take a family $A_i$ of orthogonal sets (i.e. all the elements of one set are orthogonal to all the elements of any other set) and take an element that is a mixture of elements from that family, then the fraction capacity will be additive over the lattice formed by closing the family under union and intersection. In quantum mechanics, this would correspond to all probability distributions that are outcomes of the projective measurement that distinguishes between $A_i$. Part of the task, then, will be finding a set of necessary and sufficient conditions under which the fraction capacity for a particular ensemble is additive over a particular lattice of subsets

What is important here is that
